---
title: "Tutorial: building train and test sets with the same characteristics"
header-includes: \usepackage{booktabs}
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    number_sections: yes

vignette: >
  %\VignetteIndexEntry{"Tutorial: building train and test sets with the same characteristics"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r results='hide', message=FALSE, warning=FALSE, echo=FALSE}
set.seed(2)
library(dataPreparation)
library(data.table)
library(knitr)
library(kableExtra)
library(pander)
options(knitr.table.format = "html")
Sys.setlocale("LC_TIME", "C")
```

# Introduction

## Purpose of this vignette
This vignette is a tutorial to prepare a `train` and a `test` set using `dataPreparation` package.

In this tutorial the following points are going to be viewed:

  * Preparing a training set,
  * Applying the same preparation to a testing set,
  * Controling that train and test sets have the same shape.

Using [dataPreparation](https://CRAN.R-project.org/package=dataPreparation/index.html) package, those sets will be

This package is

  * fast (use `data.table` and exponential search)
  * RAM efficient (perform operations by reference and column-wise to avoid copying data)
  * stable (most exceptions are handled)
  * easy (since those functions are packaged and handle most of the situations)
  * verbose (log a lot)

## Data set
For this tutorial, UCI [adult](https://archive.ics.uci.edu/ml/datasets/adult) data set will be used.

The goal with this data set is to predict the income of individuals based on 14 variables.

Let's have a look to the data set:
```{r comment="#",  null_prefix=TRUE}
data("adult")
print(head(adult, n = 4))
```



# Preparing data
## Spliting Train and test
To avoid introducing a bias in `test` using train-data, the train-test split should be performed before (most) data preparation steps.

To simulate a train and test set we are going to split randomly this data set into 80% train and 20% test.

```{r comment="#",  null_prefix=TRUE}
# Random sample indexes
train_index <- sample(1:nrow(adult), 0.8 * nrow(adult))
test_index <- setdiff(1:nrow(adult), train_index)

# Build X_train, y_train, X_test, y_test
X_train <- adult[train_index, -15]
y_train <- adult[train_index, "income"]

X_test <- adult[test_index, -15]
y_test <- adult[test_index, "income"]
```


## Filter useless variables

The first thing to do, in order to make computation fast, would be to filter useless variables:

- Constant variables
- Variables that are in double (for example col1 == col2)
- Variables that are exact bijections (for example col1 = A, B, B, A and col2 = 1, 2, 2, 1)

Let's id them:
```{r warning = FALSE, comment="#",  null_prefix=TRUE}
constant_cols <- which_are_constant(adult)
double_cols <- which_are_in_double(adult)
bijections_cols <- which_are_bijection(adult)
```

We only found, one bijection: variable `education_num` which is an index for variable `education`. Let's drop it:


```{r warning = FALSE, comment="#",  null_prefix=TRUE}
X_train$education_num = NULL
X_test$education_num = NULL
```


## Scaling
Most machine learning algorithm rather handle scaled data instead of unscaled data.

To perform scaling (meaning setting mean to 0 and standard deviation to 1), function `fast_scale` is available.

Since it is highly recommended to apply same scaling on train and test, you should compute the scales first using the function `build_scales`:

```{r warning = FALSE, comment="#",  null_prefix=TRUE}
scales <- build_scales(data_set = X_train, cols = c("capital_gain", "capital_loss"), verbose = TRUE)
print(scales)
```
As one can see, those to columns have very different mean and standard deviation. Let's apply scaling on those:

```{r warning = FALSE, comment="#",  null_prefix=TRUE}
X_train <- fast_scale(data_set = X_train, scales = scales, verbose = TRUE)
X_test <- fast_scale(data_set = X_test, scales = scales, verbose = TRUE)
```

And now let's have a look at the result:
```{r warning = FALSE, comment="#",  null_prefix=TRUE}
print(head(X_train[, c("capital_gain", "capital_loss")]))
```

## Discretization

One might want to discretize the variable age, either using an equal freq/width method, or some hand-written bis.

To compute equal freq bins, `build_bins` is available:
```{r warning = FALSE, comment="#",  null_prefix=TRUE}
bins <- build_bins(data_set = X_train, cols = "age", n_bins = 10, type = "equal_freq")
print(bins)
```

To make it easy to use, in this package:

- `data_set` will always denote the data.table on which you want to perform something.
- `cols` will always denote the columns on which you want to apply the function. It could also be set to "auto" to apply it on all relevant columns.
- Some spefic argument could be needed and will be presented in the documentation of each functions.


Let's apply our own bins:
```{r warning = FALSE, comment="#",  null_prefix=TRUE}
X_train <- fast_discretization(data_set = X_train, bins = list(age = c(0, 18, 25, 45, 62, +Inf)))
X_test <- fast_discretization(data_set = X_test, bins = list(age = c(0, 18, 25, 45, 62, +Inf)))
```

Here bins have been defined to compute groups :

- 0 to 18
- 18 to 25
- 25 to 45
- 45 to 62
- Over 62.

Let's control it:
```{r warning = FALSE, comment="#",  null_prefix=TRUE}
print(table(X_train$age))
```

## Encoding categorical
One thing to do when you are using some machine learning algorithm such as a logistic regression or a neural network is to encode factor variables. One way to do that is to perform one-hot-encoding. For examples:

ID | var
---|----
1  | A
2  | B
3  | C
4  | C

Would become:

ID | var.A | var.B | var.C
---|-------|-------|-------
1  | 1     | 0     | 0
2  | 0     | 1     | 0
3  | 0     | 0     | 1
4  | 0     | 0     | 1

To perform it, one could use `dataPreparation::one_hot_encoder` which uses data.table power to do it in a fast and RAM efficient way.
Since it is important to have the same columns in train and test first, one will compute the encoding:
```{r warning = FALSE, comment="#",  null_prefix=TRUE}
encoding <- build_encoding(data_set = X_train, cols = "auto", verbose = TRUE)
```

The argument cols = "auto" means that build_encoding will automatically select all columns that are either character or factor to prepare encoding.

And then one can apply them to both tables:

```{r warning = FALSE, comment="#",  null_prefix=TRUE}
X_train <- one_hot_encoder(data_set = X_train, encoding = encoding, drop = TRUE, verbose = TRUE)
X_test <- one_hot_encoder(data_set = X_test, encoding = encoding, drop = TRUE, verbose = TRUE)
```

This function is called the following way:

- data_set = X_train: means that it will perform transformation on X_train
- encoding = encoding: means that we use previously built encoding
- drop = TRUE: means that it will drop original columns
- verbose = TRUE: means that it will log to tell you what it is doing.

Even if it's not kept in the log, a progress bar has been created to see if the functions is running and how fast. This progress bar is available in most functions from this package. It can be really helpfull when you are handling really large data sets.


Let's check the dimensions of X:
```{r warning = FALSE, comment="#",  null_prefix=TRUE}
print("Dimensions of X_train: ")
print(dim(X_train))
print("Dimensions of X_test: ")
print(dim(X_test))
```


## Filtering variables
Since a lot of columns have been created, a filtering could be relevant:

```{r warning = FALSE, comment="#",  null_prefix=TRUE}
bijections <- which_are_bijection(data_set = X_train, verbose = TRUE)
print(names(X_train)[bijections])
```

Thanks to optimisations (such as exponential search), it takes only 7s to compare 111 columns together and identify bijections. Without those optimisations it can be minutes long.

`which_are_bijection` found that column sex.Male is a bijection of column sex.Female.

Let's drop one of them:

```{r warning = FALSE, comment="#",  null_prefix=TRUE}
set(X_train, NULL, names(X_train)[bijections], NULL)
set(X_test, NULL, names(X_train)[bijections], NULL)
```


# Controling shape
Last but not least, it is very important to make sure that `train` and `test` sets have the same shape (for example the same columns).

To make sure of that one could perform the following function:
```{r warning = FALSE, comment="#",  null_prefix=TRUE}
X_test <- same_shape(X_test, reference_set = X_test, verbose = TRUE)
```

No warning have been raised it's all is ok.



# Conclusion
We presented some of the functions of *dataPreparation* package. There are a few more available, plus they have some parameters to make their use easier. So if you liked it, please go check the package documentation (by installing it or on [CRAN]( https://CRAN.R-project.org/package=dataPreparation/dataPreparation.pdf))


We hope that this package is helpful, that it helped you prepare your data in a faster way.


If you would like to give us some feedback, report some issues, add some features to this package, please tell us on [GitHub](https://github.com/ELToulemonde/dataPreparation/issues). Also if you want to contribute, please don't hesitate to contact us.
